Learning CNN filters from user-drawn image makers for coconut-tree image classification

### 2020-06-13 main.tex
Updates from Overleaf

\usepackage[pdftex]{graphicx}
\ifCLASSOPTIONcompsoc
  \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
\else
  \usepackage[caption=false,font=footnotesize]{subfig}
\fi
\usepackage{comment}
\usepackage{float}
\usepackage{mathtools}
\usepackage{amsfonts}
\usepackage{bm}
\usepackage{pgf,tikz,pgfplots}
\usetikzlibrary{calc}

@@

In order to reduce the need for large annotated training sets by exploiting the user knowledge, we present a method, called FLIM (\emph{Feature Learning from Image Markers}), that needs a minimal set of images to learn the filters of each convolutional layer. The user selects the number of convolutional layers, their filter sizes, a few training images, and draws markers in regions of the input images that best discriminate the classes. \textcolor{blue}{Classes may appear in distinct clusters of a local feature space defined by patches extracted from marker pixels at each convolutional layer's input.} FLIM defines the weights of the filters as the centroids of the clusters. Such filters are expected to activate discriminant regions in the given convolutional layer's output, and the process repeats in a layer-by-layer fashion. The architecture of the CNN could also be optimized by observing its results on a small validation set or based on data visualization techniques~\cite{rauber2016visualizing}.

We demonstrate the advantages of this interactive technique on binary image classification of coconut-tree images against  VGG-16~\cite{simonyan2014very}. First, the user can better understand and control the training process by observing the effectiveness of marker selection.  Second, very few images per class (e.g., less than five) seem to be enough. \textcolor{red}{Third, by eliminating the optimization of the convolutional layers by backpropagation, FLIM can reduce the number of annotated images to train the fully connected layers.} Fourth, FLIM is application-independent, and so it might be useful in other image classification problems.

@@

\textcolor{blue}{A well-trained CNN model should detect image regions that best discriminate classes as positive activations, improve class separation by eliminating negative activations, and aggregate the resulting activations locally (pooling) and globally (flattening) into a highly dimensional and sparse feature space, in which even a linear classifier should separate the classes with reasonable success.} Nevertheless, it is common to assume that the classes are at least piecewise linearly separable and apply fully connected layers -- a Multi-Layer Perceptron (MLP) classifier -- for feature space reduction and classification. The training of a CNN usually relies on weight optimization using backpropagation, which requires a higher number of annotated images as deep as the model. In FLIM, we eliminate the need for backpropagation to train convolutional layers by finding a set ${\cal F}$ of filters that can detect discriminant regions from all classes.

The user selects a few training images to compose a very small dataset $\D$ and draws labeled markers in image regions that best discriminate the classes. The convolution between an image $I$ and a filter $F\in {\cal F}$ generates positive activations in regions whose local patterns are detected by $F$. We wish then to estimate the weights of $F$ such that those local patterns are characteristic of one given class. For sake of clarity, one must interpret the convolution operation at a pixel $p$ of $I$ as the inner product $\vect(P_I(p)) \cdot \vect(F)$, where $P_I(p)$ is a $k\times k$ patch with $m$ channels around $p$, $F$ is a filter with shape $k\times k \times m$, and $\vect$ is the vectorization operation. A filter $F$ can discriminate a class $i \in \{1,2,\ldots,c\}$ among $c$ classes when it generates positive results for local patterns $\vect(P_I(p))$ of class $i$ and negative results for patterns from other classes. That is, $\vect(F)$ is the normal vector of a hyperplane in $\R^{k\times k\times m}$ which detects the patterns of class $i$ at its positive side. We wish to estimate ${\cal F}$ such that its filters will detect patterns from all classes in different positions $p$.

\textcolor{blue}{For a problem with $c$ classes where $\lambda(p)=i\in \{1,2,\ldots,c\}$ is the label of a marker pixel $p$ from class $i$, let ${\cal M}_I$ be a set of marker pixels drawn in image $I\in {\cal D}$ and $P_I(p)$ be the respective patch around a pixel $p\in {\cal M}_I$. Let ${\cal P}$ be the set of all patches around marker pixels from all images $I\in {\cal D}$, with representative examples  $\vect(P_I(p))$ from all the $c$ classes. We define the set of patches from class $i$ as
\begin{eqnarray}
{\cal P}_i & = & \bigcup_{I\in {\cal D},p\in {\cal M}_I, \lambda(p) = i}{P_I(p)}.
\label{eq:pset}
\end{eqnarray}
A clustering operation on ${\cal P}$ creates groups with similar local patterns. Ideally, each group should contain patterns of class $c$. However, some classes might share similar patterns. We should at least expect that their patterns be detected in different image regions. The groups must be shifted to the origin of $\R^{k\times k\times m}$ by subtracting the mean value of each patch in ${\cal P}$, where $\mathcal{P} = \bigcup_{i \in \{1, 2, \ldots, c\}}{\mathcal{P}_i}$. Additionally, the standard deviation of each patch is computed and used for standardization. This operation allows batch normalization of image sets using the mean and standard deviation of the patches from the image markers -- i.e.,  \emph{a marker-based batch normalization}. The centroid of each cluster defines the weights of each filter $F\in {\cal F}$ and we force norm $\|\vect(F)\|=1$ to avoid preferences among them. The centralization for filter definition is paramount to eliminate activations in regions from other classes, whose patterns fall in the negative side of the hyperplane with normal $\vect(F)$. Figure~\ref{fig:filter} illustrates three groups with color-coded samples from two classes in a hypothetical 2D feature space. The marker-based batch normalization and convolution with filters obtained as centroids of the three groups creates a new 3D feature space (Figure~\ref{fig:filter}b), in which the classes can be more easily detected by two filters.}

@@

Each convolutional layer is trained individually, one layer at a time, to find its filter set ${\cal F}$. The number of filters per layer depends on the clustering technique. We use $K$-means, then the user must specify the number $K=|{\cal F}|$ of filters. After the convolution operation, we apply the ReLU function to eliminate negative activations and the max-pooling operation to aggregate local information. Note that, apart from the initial marker selection, the training process is automatic. We have preserved the convolutional layers' output dimensions and the marker pixels so we can find the filters of layer $L+1$ from the output of layer $L$. Figure~\ref{fig:tree} shows one example of a simple network projected by FLIM for the experiments of the next section.

@@

  \begin{tikzpicture}
    [
      grow                    = down,
      sibling distance        = 5em,
      level distance          = 3em,
      edge from parent/.style = {draw=black, -latex},
      every node/.style       = {treenode},
      sloped
    ]
    \node [treenode, fill=tealDeer] (v0) {MarkerBased BatchNorm}
    child {
        node [treenode, fill=tealDeer] {Conv$(60 \times 7 \times 7)$}
        child {
          node [treenode, fill=tealDeer] {ReLU} 
          child {
            node [treenode, fill=tealDeer] {MaxPool$(3\times3)$}
            child {
              node [treenode, fill=tealDeer] {BatchNorm}
              child {
                node [treenode, fill=navyPurple] {Classifier}
                edge from parent
              }
            }
          } 
        }
      };

  \end{tikzpicture}
  \end{center}
  \caption{Example of a network with a single convolutional layer followed by a classifier, as projected by FLIM and used for the experiments in this work.}
  \label{fig:tree}
\end{figure}

@@

We use a dataset from~\cite{8899005}, which contains regions of aerial images with and without coconut trees from the Kingdom of Tonga, as acquired in October 2017. Each region should be classified as containing one or none coconut tree, but part of coconut trees from adjacent regions might appears near the region's border. The Humanitarian OpenStreetMap community annotated the regions. The dataset consists of 13587 regions with $90 \times 90$ pixels of size $8\times 8$  $cm^2$, being 10268 and 3319 regions annotated as containing and not containing coconut trees, respectively.

In order to assess the impact of FLIM in reducing the need for annotated samples, we randomly selected 100 regions from each class for training and left the remaining 11387 regions for testing. We also repeated this procedure three times in order to compute mean and standard deviation of the results in precision, recall, and f-score. By vectorizing and projecting the 200 training images of each split on 2D~\cite{maaten2008visualizing}, the user identified cohesive groups of images from each class and selected two representative images from each class to draw markers. We neither evaluated markers from a higher number of images nor multiple network architectures trained by FLIM. We just trained by FLIM the simplest network with a single convolutional layer from markers drawn in four images only.

For comparative analysis, we select the popular VGG-16~\cite{simonyan2014very} since it is a deep neural network that usually requires a much higher number  of annotated samples than 200 to be trained from the scratch. It is also a very competitive CNN when pretrained in the ImageNet dataset. Therefore, as baselines, we use three solutions based on VGG-16: (1) VGG is a network trained from the scratch, using the Xavier's weight-initialization approach, (2) VGG-FT is the network pretrained in the ImageNet dataset and fine-tuned in the coconut-tree training set, and (3) VGG-FE is the feature extractor obtained at the output of the last convolutional layer (layer 13th) of VGG-16 pretrained in the ImageNet dataset. VGG and VGG-FT use a MLP classifier as part of the network and we use the linear SVM-OVO classifier from sklearn~\cite{fan2008liblinear} with $C=0.01$ for VGG-FE.

In order to evaluate the quality of our feature learning approach, we first compare the features produced by a single-layer network trained by FLIM with those of the last convolutional layer (layer 13th) of the VGG-16, pretrained on the ImageNet dataset. For this comparison, we use  (VGG-FE). Table~\ref{tab:results} shows the result of this comparison on the first two rows.

We compare different solutions based on FLIM and In order to evaluate FLIM metho, we select VGG-16 as a baseline CNN. VGG-16

As in reality, labeling images is a costly and time-consuming process, and then we assume that we only knew the correct labels of a small set of samples (one hundred images for each class, totaling two hundred images). That way, we would have only those two hundred images to train any models that we would use to try to solve the problem.

To reduce the impact of chance on the results, we randomly selected three sets of 200 images for training and three sets of 11387 images for testing. We wanted to represent a scenario where the network designer did not have much effort to annotate all these images. From each of these sets of training images, we selected four images using the 2D space projection of the 200 images using the t-SNE. We tried to select images that were part of more cohesive groups. Thus, these images are more likely to have characteristics that represent their group.

\begin{figure}[!t]
    \centering
    \subfloat[\label{fig:markers1}]{\includegraphics[width=.3\linewidth]{figures/markers1}
    }
    ~
    \subfloat[\label{fig:markers2}]{\includegraphics[width=.3\linewidth]{figures/markers2}
    }
    \\
    \subfloat[\label{fig:markers3}]{\includegraphics[width=.3\linewidth]{figures/markers3}
    }
    ~
    \subfloat[\label{fig:markers4}]{\includegraphics[width=.3\linewidth]{figures/markers4}
    }
    \caption{Markers used for training. (a) Markers on an image which contains a coconut tree. (b) Markers on an image which does not contain a coconut tree. (c) Markers on an image which does not contain a coconut tree. (d) Markers on an image which does not contain a coconut tree.}
    \label{fig:markers}
\end{figure}

We placed markers on these images in regions that we consider important to identify the coconut and non-coconut (more diverse) classes. In Figure \ref{fig:markers} we can the four images of one the splits. With the markers, we created a convolutional layer with filters of dimensions $7 \times 7$ using the method described in Section \ref{sec:method}. The network architecture can be seen in the Figure. We used a classifier similar to that of VGG-16. We used K-means to cluster the coconut tree patches in 30 clusters and the non-coconut tree patches in also 30 clusters to find the convolutional layer filter.

@@

    FLIM & $0.863 \pm 0.002$ & $0.849 \pm 0.005$ & $0.854 \pm 0.004$\\
    FLIM-FT & $0.845 \pm 0.003$ & $0.825\pm 0.006$ & $0.832 \pm 0.005$ \\

@@

  \hline
  \end{tabular}
  \end{center}
  \caption{Results.}
  \label{tab:results}
\end{table}

We tested three scenarios: (I) using out method to learn the feature extractor weights and the classifier with backpropagation; (II) using our method as initialization and train the whole model with backpropagation; (III) using SVM Linear as classification. For VGG-16, we tested two scenarios: (IV) initializing with Xavier initialization and training from scratch; (V) using ImageNet weights and training with the training set images. The results can be seen in Table \ref{tab:results}. As the experiments were repeated three times, we reported the mean and standard deviation of each the metrics in the test set.

\begin{figure}
  \centering
  \subfloat{\includegraphics[width=.3\linewidth]{figures/qualitative/ours/upper-coco-108.png}
  }
  ~
  \subfloat{\includegraphics[width=.3\linewidth]{figures/qualitative/ours/upper-coco-2473.png}
  }
  ~
  \subfloat{\includegraphics[width=.3\linewidth]{figures/qualitative/ours/upper-non_coco-506.png}
  }
  \\
  \subfloat{\includegraphics[width=.3\linewidth]{figures/qualitative/vgg/lower-coco-1098.png}
  }
  ~
  \subfloat{\includegraphics[width=.3\linewidth]{figures/qualitative/vgg/lower-coco-2400.png}
  }
  ~
  \subfloat{\includegraphics[width=.3\linewidth]{figures/qualitative/vgg/lower-non_coco-105.png}
  }
  \caption{Misclassified images. The first row has images misclassified by the network trained by our method, and the second row has images misclassified by VGG-16. The two first images of a row have coconut trees.}
  \label{fig:ex-classification}
\end{figure}

Our method achieved better precision than conventionally trained VGG, and the difference between precision and recall is much smaller. The Linear SVM did not prove to be a good classifier like the MLP, but, on average, it still got a little better than the VGG-16. Using our method as initialization and training the entire network with backpropagation did not improve the result; this implies that the filters we created are already good filters given the training set. Our feature extractor training method was able to learn good filters from a minimal set of images. VGG has many parameters to learn, so it is susceptible to overfitting with such a small training set. We do not use any technique to handle small data sets to train VGG because we intended to verify our approach viability, and these techniques could also be used with our method. In Figure \ref{fig:ex-classification} we can see examples of misclassified images. In these images, coconut trees appear in different angles, sizes, and shapes, or the boundaries between the coconut tree and the background are tenuous, making it more difficult to identify them. In turn, images that do not contain coconut trees, contain trees that resemble their shape. The network designer could put markers in those images in other to improve the feature extractor.

\begin{figure}
  \centering
  \subfloat[\label{fig:vis-input}]{\includegraphics[width=.3\linewidth]{figures/input}
  }
  ~
  \subfloat[\label{fig:vis-feature-extractor}]{\includegraphics[width=.3\linewidth]{figures/feature_extractor}
  }
  ~
  \subfloat[]{\includegraphics[width=.3\linewidth]{figures/classifier}
  \label{fig:vis-classifier}}
  \caption{This image shows visualizations of the input images and the intermediate layer's outputs. Red points represent background images, and green points represent images that have a coconut tree. The projection from high dimensionality space to 2D was made using t-SNE. The original image shows much overlap between the classes; after applying the feature extractor with only one convolutional layer, we can see that the overlapping decreases considerably. Before the decision layer, the last layer put most of the images from one class on one side and the other on the other side. (a) Visualization of original image feature space. (b) Visualization of the feature extractor's outputs. (c) Visualization of outputs from the last classifier hidden layer.}
\end{figure}

As motivated by Rauber et al. \cite{rauber2016visualizing}, we made projections of three stages of the network using t-SNE \cite{maaten2008visualizing} to understand how the feature extractor trained with our method transforms the image spaces. In Figure \ref{fig:vis-input}, we can see the projection of the original images (in the LAB color space). Red points are images that contain coconut trees, and green points are images that do not contain coconut trees. In the projection in the image of Figure \ref{fig:vis-input}, we see that the images of coconut trees and non-coconut trees are almost entirely overlapping. Coconut tree images are more frequent in some regions but are very dispersed. In the image of Figure \ref{fig:vis-feature-extractor}, we see the projection of the feature extractor output. The coconut tree images are more concentrated, forming some clusters, and it is possible to notice that there is much less overlap. Finally, we see in Figure \ref{fig:vis-classifier} the output of the previous layer of the decision layer of the classifier. The classifier managed to organize conquest and non-coconut images in different regions of space, although there is overlap as the model failed to achieve 100\% accuracy.

\section{Conclusions}

We present a way to train a feature extractor without using backpropagation capable of learning good filters from a minimal set of images. To evaluate the method, we proposed experiments to classify images that contain achievements. Our approach has achieved competitive results with state of the art.

\bibliographystyle{IEEEtran}
\bibliography{bibliography}

>>>>

\usepackage[pdftex]{graphicx}
\ifCLASSOPTIONcompsoc
  \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
\else
  \usepackage[caption=false,font=footnotesize]{subfig}
\fi
\usepackage{comment}
\usepackage{float}
\usepackage{mathtools}
\usepackage{amsfonts}
\usepackage{bm}
\usepackage{siunitx}
\usepackage{pgf,tikz,pgfplots}
\usetikzlibrary{calc}

@@

In order to reduce the need for large annotated training sets by exploiting the user knowledge, we present a method, called FLIM (\emph{Feature Learning from Image Markers}), that needs a minimal set of images to learn the filters of each convolutional layer. The user selects the number of convolutional layers, their filter sizes, a few training images, and draws markers in regions of the input images that best discriminate the classes. Classes may appear in distinct clusters of a local feature space defined by patches extracted from marker pixels at each convolutional layer's input. FLIM defines the weights of the filters as the centroids of the clusters. Such filters are expected to activate discriminant regions in the given convolutional layer's output, and the process repeats in a layer-by-layer fashion. The architecture of the CNN could also be optimized by observing its results on a small validation set or based on data visualization techniques~\cite{rauber2016visualizing}.

We demonstrate the advantages of this interactive technique on binary image classification of coconut-tree images against  VGG-16~\cite{simonyan2014very}. First, the user can better understand and control the training process by observing the effectiveness of marker selection.  Second, very few images per class (e.g., less than five) seem to be enough. Third, by eliminating the optimization of the convolutional layers by backpropagation, FLIM can reduce the number of annotated images to train the fully connected layers. Fourth, FLIM is application-independent, and so it might be useful in other image classification problems.

@@

A well-trained CNN model should detect image regions that best discriminate classes as positive activations, improve class separation by eliminating negative activations, and aggregate the resulting activations locally (pooling) and globally (flattening) into a highly dimensional and sparse feature space, in which even a linear classifier should separate the classes with reasonable success. Nevertheless, it is common to assume that the classes are at least piecewise linearly separable and apply fully connected layers -- a Multi-Layer Perceptron (MLP) classifier -- for feature space reduction and classification. The training of a CNN usually relies on weight optimization using backpropagation, which requires a higher number of annotated images as deep as the model. In FLIM, we eliminate the need for backpropagation to train convolutional layers by finding a set ${\cal F}$ of filters that can detect discriminant regions from all classes.

The user selects a few training images to compose a very small dataset $\D$ and draws labeled markers in image regions that best discriminate the classes. The convolution between an image $I$ and a filter $F\in {\cal F}$ generates positive activations in regions whose local patterns are detected by $F$. We wish then to estimate the weights of $F$ such that those local patterns are characteristic of one given class. For sake of clarity, one must interpret the convolution operation at a pixel $p$ of $I$ as the inner product $\vect(P_I(p)) \cdot \vect(F)$, where $P_I(p)$ is a $k\times k$ patch with $m$ channels around $p$, $F$ is a filter with shape $k\times k \times m$, and $\vect$ is the vectorization operation. A filter $F$ can discriminate a class $i \in \{1,2,\ldots,c\}$ among $c$ classes when it generates positive results for local patterns $\vect(P_I(p))$ of class $i$ and negative results for patterns from other classes. That is, $\vect(F)$ is the normal vector of a hyperplane in $\R^{k\times k\times m}$, which detects the patterns of class $i$ at its positive side. We wish to estimate ${\cal F}$ such that its filters will detect patterns from all classes in different positions $p$.

For a problem with $c$ classes where $\lambda(p)=i\in \{1,2,\ldots,c\}$ is the label of a marker pixel $p$ from class $i$, let ${\cal M}_I$ be a set of marker pixels drawn in image $I\in {\cal D}$ and $P_I(p)$ be the respective patch around a pixel $p\in {\cal M}_I$. Let ${\cal P}_i$ be the set of all patches around marker pixels from all images $I\in {\cal D}$, with representative examples  $\vect(P_I(p))$ from class $i\in \{1,2,\ldots,c\}$. 
\begin{eqnarray}
{\cal P}_i & = & \bigcup_{I\in {\cal D},p\in {\cal M}_I, \lambda(p) = i}{P_I(p)}.
\label{eq:pset}
\end{eqnarray}
A clustering operation on each set ${\cal P}_i$, $i\in [1,c]$, guarantees groups with similar local patterns for each class. The groups must be shifted to the origin of $\R^{k\times k\times m}$ by subtracting the mean value of all patches, set $\mathcal{P} = \bigcup_{i \in \{1, 2, \ldots, c\}}{\mathcal{P}_i}$. Additionally, the standard deviation of all patches is computed and used for standardization. This operation allows batch normalization of image sets using the mean and standard deviation of the patches from the image markers -- i.e.,  \emph{a marker-based batch normalization}. The centroid of each cluster defines the weights of each filter $F\in {\cal F}$, and we force norm $\|\vect(F)\|=1$ to avoid preferences among them. The centralization for filter definition is paramount to eliminate activations in regions from other classes, whose patterns fall in the negative side of the hyperplane with normal $\vect(F)$. Figure~\ref{fig:filter} illustrates three groups with color-coded samples from two classes in a hypothetical 2D feature space. The marker-based batch normalization and convolution with filters obtained as centroids of the three groups creates a new 3D feature space (Figure~\ref{fig:filter}b), in which the classes can be more easily detected by two filters.

@@

Each convolutional layer is trained individually, one layer at a time, to find its filter set ${\cal F}$. The number of filters per layer depends on the clustering technique. We use $K$-means, and then the user must specify the number $K=|{\cal F}|$ of filters. After the convolution operation, we apply the ReLU function to eliminate negative activations and the max-pooling operation to aggregate local information. Note that, apart from the initial marker selection, the training process is automatic. We have preserved the convolutional layers' output dimensions and the marker pixels so we can find the filters of layer $L+1$ from the output of layer $L$. Figure~\ref{fig:arch} shows one example of a simple network projected by FLIM for the experiments of the next section.

@@

  \begin{tikzpicture}
    [
      grow                    = down,
      sibling distance        = 5em,
      level distance          = 8mm,
      edge from parent/.style = {draw=black, -latex},
      every node/.style       = {treenode},
      sloped
    ]
    \node [treenode, fill=tealDeer] (v0) {MarkerBased BatchNorm}
    child {
        node [treenode, fill=tealDeer] {Conv$(60 \times 7 \times 7)$}
        child {
          node [treenode, fill=tealDeer] {ReLU} 
          child {
            node [treenode, fill=tealDeer] {MaxPool$(3\times3)$}
            child {
              node [treenode, fill=tealDeer] {BatchNorm}
              child {
                node [treenode, fill=navyPurple] {Classifier}
                edge from parent
              }
            }
          } 
        }
      };

  \end{tikzpicture}
  \end{center}
  \caption{Example of a network with a single convolutional layer followed by a classifier, as projected by FLIM and used for the experiments in this work.}
  \label{fig:arch}
\end{figure}

@@

We used a dataset from~\cite{8899005}, which contains regions of aerial images with and without coconut trees from the Kingdom of Tonga, as acquired in October 2017~\footnote{The images are available in \url{blog.werobotics.org/2018/01/11/open-ai-challenge-2}.}. Each region should be classified as containing one or none coconut tree, but part of coconut trees from adjacent regions might appears near the region's border. The Humanitarian OpenStreetMap community annotated the regions. The dataset consists of 13587 regions with $90 \times 90$ pixels with resolution under $10$\si{\square\centi\metre}, being 10268 and 3319 regions annotated as containing and not containing coconut trees, respectively. The region images are converted to the LAB color space, with each channel normalized within $[0,1]$.

To assess the impact of FLIM in reducing the need for annotated samples, we randomly selected 100 regions from each class for training and left the remaining 11387 regions for testing. We repeated this procedure three times to compute the mean and the standard deviation of the results in precision, recall, and f-score. By vectorizing and projecting the 200 training images of each split on 2D with t-SNE~\cite{maaten2008visualizing}, we identified cohesive groups of images from each class and selected two representative images from each class to draw markers.  The selected images from one split can be seen in Figure \ref{fig:markers}. We placed markers on these images in regions that we thought to better distinguish between coconut and non-coconut classes.

We neither evaluated markers from a higher number of images nor multiple network architectures trained by FLIM. We just trained the single-convolutional-layer network of Figure \ref{fig:arch} from markers drawn in two images per class.
The convolutional layer has $60$ filters with dimension $7 \times 7$ and 3 color channels. It includes marker-based normalization before convolution, ReLu activation, and a max-pooling operation with a window of dimension $3 \times 3$. Since it is the single and last convolutional layer, we applied a stride of $4$ in max-pooling and batch normalization to create an input as close as possible to the one of the MLP classifier used in the popular VGG-16~\cite{simonyan2014very}.

We selected VGG-16 for comparative analysis since it usually requires many more samples than 200 to be trained from scratch. It is also a very competitive CNN when pretrained in the ImageNet dataset, working as an alternative solution for small training sets. Therefore, as baselines, we used three solutions based on VGG-16: (1) VGG is trained from scratch, (2) VGG-FT is pretrained in the ImageNet dataset and fine-tuned in the coconut-tree training set, and (3) VGG-FE is the feature extractor of VGG-16 (the output of the last convolutional layer) pretrained in the ImageNet dataset. In this case, we used for classification the  linear SVM classifier with $C=0.01$ from sklearn~\cite{fan2008liblinear, scikit-learn}.

Table~\ref{tab:results} shows the results of three FLIM-based solutions using the SVM and MLP classifiers; "FLIM + SVM", "FLIM + MLP", and "FLIM-FT + MLP" (i.e., FLIM with fine-tuning); compared with the VGG-based solutions above. The first two rows indicate that FLIM  can provide a more effective feature extractor using a single convolutional layer than VGG-FE, which contains 13 convolutional layers pretrained on the ImageNet. The results of "FLIM + MLP" and "FLIM-FT + MLP" show that FLIM  may dismiss fine tuning by providing a more effective solution. By comparing "FLIM + MLP" with VGG, one can observe that FLIM reduces the number of required training images. Finally, the comparison between "FLIM + MLP" and VGG-FT demonstrates that FLIM can provide competitive results using a considerably more simplified architecture than  VGG-FT. The number of parameters of our feature extractor represents only $0.07\%$ of the number of parameter of VGG16's feature extractor.

\begin{figure}[!t]
    \centering
    \subfloat[\label{fig:markers1}]{\includegraphics[width=.3\linewidth]{figures/markers1}
    }
    ~
    \subfloat[\label{fig:markers2}]{\includegraphics[width=.3\linewidth]{figures/markers2}
    }
    \\
    \subfloat[\label{fig:markers3}]{\includegraphics[width=.3\linewidth]{figures/markers3}
    }
    ~
    \subfloat[\label{fig:markers4}]{\includegraphics[width=.3\linewidth]{figures/markers4}
    }
    \caption{Markers used for training in image regions with (first column) and without (second column) a coconut tree.}
    \label{fig:markers}
\end{figure}

@@

    FLIM + MLP & $0.863 \pm 0.002$ & $0.849 \pm 0.005$ & $0.854 \pm 0.004$\\
    FLIM-FT + MLP & $0.845 \pm 0.003$ & $0.825\pm 0.006$ & $0.832 \pm 0.005$ \\

@@

  \hline
  \end{tabular}
  \end{center}

\caption{Mean and standard deviation of precision, recall, and f-score for each method: a feature extractor with a classifier.}
  \label{tab:results}
\end{table}

\begin{figure*}
  \centering
  \subfloat{\includegraphics[width=.13\linewidth]{figures/qualitative/ours/upper-coco-108.png}
  }
  ~
  \subfloat{\includegraphics[width=.13\linewidth]{figures/qualitative/ours/upper-coco-2473.png}
  }
  ~
  \subfloat{\includegraphics[width=.13\linewidth]{figures/qualitative/ours/upper-non_coco-506.png}
  }
  ~
  \subfloat{\includegraphics[width=.13\linewidth]{figures/qualitative/vgg/lower-coco-1098.png}
  }
  ~
  \subfloat{\includegraphics[width=.13\linewidth]{figures/qualitative/vgg/lower-coco-2400.png}
  }
  ~
  \subfloat{\includegraphics[width=.13\linewidth]{figures/qualitative/vgg/lower-non_coco-105.png}
  }
  \caption{Misclassified images. The first row has images misclassified by the network trained by our method, and the second row has images misclassified by VGG-16. The two first images of a row have coconut trees.}
  \label{fig:ex-classification}
\end{figure*}

 In Figure \ref{fig:ex-classification}, we can see examples of misclassified images. In these images, coconut trees appear in different angles, sizes, and shapes, or the boundaries between the coconut tree and the background are tenuous, making it more difficult to identify them. In turn, images that do not contain coconut trees, contain trees that resemble their shape. The network designer could put markers in those images in other to improve the feature extractor.

\begin{figure}
  \centering
  \subfloat[\label{fig:vis-input}]{\includegraphics[width=.3\linewidth]{figures/input}
  }
  ~
  \subfloat[\label{fig:vis-feature-extractor}]{\includegraphics[width=.3\linewidth]{figures/feature_extractor}
  }
  ~
  \subfloat[]{\includegraphics[width=.3\linewidth]{figures/classifier}
  \label{fig:vis-classifier}}
  \caption{Projections by t-SNE of the images at (a) the input layer (original images), (b) the output of the convolutional layer of our network (Figure~\ref{fig:arch}), and (c) the output of the last hidden layer of the MLP classifier in "FLIM + MLP".}
\end{figure}

As motivated by Rauber et al. \cite{rauber2016visualizing}, we made projections of three stages of the network "FLIM + MLP" using t-SNE \cite{maaten2008visualizing} to understand how our CNN transforms the image spaces. In Figure \ref{fig:vis-input}, we can see the projection of the original images (in the LAB color space). Green points are images that contain coconut trees, and red points are images that do not contain coconut trees. The projection shows considerable overlapping between the two classes. Coconut-tree images are more frequent in some regions but they appear very dispersed. In Figure \ref{fig:vis-feature-extractor}, the projection of the output of our feature extractor shows some reduction in the dispersion of the coconut-tree images. The coconut tree images are more concentrated, forming some clusters, and it is possible to notice some overlapping reduction between classes. Finally, Figure \ref{fig:vis-classifier} shows the output of the last hidden layer of the MLP classifier. The overlapping between classes and the dispersion of coconut-tree images are considerably reduced.

\section{Conclusion}

We introduced a first feature learning technique, named FLIM, that can estimate effective filter weights for the convolutional layers of a given network architecture from user-drawn markers in a few training images. The resulting feature extractor can be used with different classifiers and, when it is used with a MLP classifier, the number of required images to train the fully connected layers is reduced. We demonstrated the advantages of FLIM over solutions based on VGG-16 for binary classification of coconut-tree images. The experiments indicated that FLIM is an effective approach that can produce considerably simplified network architectures. By involving the user in the training process, FLIM improves the understanding about CNN models and the user control over their training process.

We intend to further investigate feature learning from image markers for different applications, elaborate methodologies to optimize the network architecture, and extend FLIM to estimate weights in fully connected layers.

\section*{Acknowledgements}

The authors thank FAPESP (Proc. 2014/12236-1), CNPq (Proc. 303808/2018-7), and Unisinos (Viscarb and Visorg projects). 
\bibliographystyle{IEEEtran}
\bibliography{bibliography}
